The Markov Link Method: a nonparametric approach to combine observations from multiple experiments

### 2018-08-10 paper/paper.tex
More elaborate explanation of method

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $N^X$ and $N^Y$.  Here $N^X_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $N^Y_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.  For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses the data to produce two objects:
%
\begin{enumerate}
    \item $\hat q(y|x)$ -- a point estimate for the link parameter $q^*(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link parameter $q^*(y|x)$ with high probability
\end{enumerate}
%
To produce these objects we proceeds as follows.  Let $\tilde P(\cdot |\ell) \sim \mathrm{Dirichlet}(N^X_\ell+1)$ and $\tilde H(\cdot |\ell) \sim \mathrm{Dirichlet}(N^Y_\ell+1)$ for each $\ell$.  For any pair $h,h'$ let 
\[
D(h|h') = \sum_{\ell,y} h(y|\ell) \log \frac{h(y|\ell)}{h'(y|\ell)}
\]
denote the Kullback-Leibler divergence from $h$ to $h'$.  For each $\tilde P,\tilde H$, let $E(\tilde P,\tilde H)$ denote the minimum value of $D(\tilde H|h')$ among all $h'$ such that there exists $q$ satisfying $h'(y|\ell)=\sum_x \tilde P(y|\ell)q(y|x)$.  Let $P,H$ denote variables carrying the distribution of $\tilde P,\tilde H$ after conditioning on $E(\tilde P,\tilde H)=0$.   Let $\Theta(p,h)=\{q:\ h(y|\ell) = \sum_x p(x|\ell)q(y|x)\}$.  We then define the MLM objects as follows:
\begin{enumerate}
    \item $\hat q\triangleq\mathbb{E}[\argmin_{q\in \Theta(P,H)} \sum_{x,y} q(y|x)^2]$.
    \item The lower bound of $C_{x,y}$ is defined as the largest number $c$ such that $\min_{q\in \Theta(P,H)} q(y|x) \geq c$ with probability 97.5\%.  Likewise the upper bound is the smallest number $c$ such that $\max_{q\in \Theta(P,H)} q(y|x) \leq c$ with probability 97.5\%.
\end{enumerate} 
%
Code to compute these objects is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

Our approach arises naturally from a Bayesian point of view.  To understand how, first notice that there are several `nuisance parameters' we must consider in order to estimate the link $q^*$.  Let $p^*(x|\ell)$ denote the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$ and $h^*(y|\ell) \triangleq \sum_x p^*(x|\ell)q(y|x)$ denote the probability such a cell will have morphological type $y$.  Note that the distribution of the data $N^X,N^Y$ is entirely determined by $p^*,h^*$.   In particular, if someone changed $q^*$ without changing $p^*,h^*$ there would be no way for us to know.  If this can occur, $q^*$ is said to be `nonidentifiable.'   Therefore, instead of trying to directly estimate $q^*$ (which may be impossible to accurately estimate) we focus on three quantities which can be estimated from data:
%
\begin{align*}
q^*_{\mathrm{mid}} &\triangleq \argmin_{q\in \Theta(p^*,h^*)} \sum_{x,y} q(y|x)^2\\
q^*_{\mathrm{lo},x,y} &\triangleq \min_{q\in \Theta(p^*,h^*)} q(y|x) \\
q^*_{\mathrm{hi},x,y} &\triangleq \max_{q\in \Theta(p^*,h^*)} q(y|x) 
\end{align*}
%
Note that all three of these optimization problems have unique solutions and would be straightforward to solve if $p^*,h^*$ were known; optimization details can be found in Appendix \ref{sec:optproblems}.  Moreover, regardless of any identifiability issues for $q^*$ itself, the objects $q^*_{\mathrm{lo},x,y}$ and $q^*_{\mathrm{hi},x,y}$ are always identifiable for every $(x,y)$ and define bounds on the true value of $q^*(y|x)$.

The MLM algorithm therefore adopts a Bayesian point of view to estimate $q^*_{\mathrm{mid}}$ and create a credible interval for $q^*_{\mathrm{lo},x,y},q^*_{\mathrm{hi},x,y}$.  We begin with uniform prior beliefs for the parameters of $p^*,h^*$.  That is, our prior is that $p^*(\cdot|\ell),h^*(\cdot|\ell)$ carry $\mathrm{Dirichlet}(1)$ distributions for each $\ell$.  Following the Bayesian philosophy, we would like to incorporate any additional knowledge through conditioning.  We have two important pieces of knowledge we would like to incorporate: that the MLM assumption holds (i.e.\ there is some $q^*$ so that $\sum_x p^*(x|\ell)q^*(y|x)=h^*(y|\ell)$) and we have observed the data $N^X,N^Y$.  By conditioning on this knowledge, we obtain the posterior distribution carried by the variables $P,H$ described above.  In terms of these variables, the MLM estimator $\hat q$ is simply the posterior mean for $q^*_{\mathrm{mid}}$ and the MLM credible interval $C_{x,y}$ is an interval which contains $q^*_{\mathrm{lo},x,y},q^*_{\mathrm{hi},x,y}$ with high probability.

In practice, we were not able to find a way to compute the credible interval and posterior means exactly.  It seems difficult to even sample the conditional random variables $P,H$.  Common approaches to this type of problem involve Markov-Chain Monte Carlo and Variational Inference, but we were unable to make these approaches work in practice.  It seems nontrivial to work with the condition $E(p,h)=0$ that formalizes the MLM assumption.  We instead take a somewhat na\"ive approach.  We start with $\tilde P,\tilde H$, which carry the distribution of $p^*,h^*$ conditioned on the observed data (but not conditioned on the fact that the MLM assumption holds).  We then simply project $\tilde H$ to the space of parameters which are consistent with the MLM assumption.  That is, we take $\hat H$ to be the solution of the minimization problem $\min_{h'}D(\tilde H|h')$, subject to the constraint that $E(\tilde P,h')=0$.  Optimization details can be found in Appendix \ref{sec:optproblems}.  We hope that the distribution of $\tilde P,\hat H$ will approximate the desired posterior $P,H$.  We then use samples from this distribution to compute what we need.  The implications of this approximation are difficult to reason about; we hope further study will clarify when this is a reasonable procedure.  In the mean time, we content ourselves that it is a practical solution to a difficult problem.

A frequentist point of view might provide a valuable counterpoint to our Bayesian analysis.  However, we were unable to produce effective confidence intervals for the link parameters.  There are at least two difficulties: the link is unidentifiable and the true parameter values may lie on the edge of the parameter space.  These challenges make traditional appproaches based on asymptotic normality difficult to apply.  Perhaps future work could uncover a way to overcome these challenges.

@@

\section{Simulations}

@@

We used simulations to test the two parts of the MLM: the point estimators $\hat q(y|x)$ and credible intervals $C_{x,y}$.  We would ordinarily hope for two things:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat q(y|x)$ should converge to the true link $q^*(y|x)$ for each $x,y$.
    \item[Conservative coverage] $q^*(y|x) \in C_{x,y}$ with high probability.
\end{description}
Our simulations yielded good news and bad news.  The good news: the MLM generally achieves conservative coverage.  In many cases, the MLM also achieves estimator convergence.  The bad news: in many cases the MLM fails to achieve estimator convergence.  Indeed, we can show that the identifiability issues block any possibility of success in some cases; we detail this problem in Appendix \ref{sec:casestudies}.  The following two simulations give a good overview of the relevant issues:

@@

The MLM appears to provide reasonable values in simulation and gives useful insight to real data.  Applied to neuronal cell-type data, the MLM clarified how two different cell-type classification systems are related.  We saw that some aspects of the two systems seem well-calibrated, but others we are less sure about.  The nature of the variability in the credible intervals suggested directions for experiments which would further refine our understanding.

@@

More generally, it is clear that good assumptions can help us get real insight for tough problems. Even if these assumptions are not sufficient to allow us to perfectly identify our object of interest, credible regions can be constructed.  By probing these regions carefully, we can learn what the data actually has to say and what experiments will help us learn more.

@@

This theorem is certainly a step in the right direction, but we emphasize that it does require some assumptions.  It is our opinion that the linear independence requirement is fairly mild (if it is not met then subpopulations can simply be merged together).  However, the positivity requirement is quite troubling.  In many cases of interest the true link $q^*$ has genuine zeros: pairs of measurements between the two tools which are fundamentally incompatible.  In this case such a theorem cannot be applied.

However, it may be that the above theorem's requirement ($q(x|y)>c>0$) is much stronger than is necessary.  A precise understanding of the discontinuity example above has remained elusive.  Better understanding could lead to a much more accurate and practical estimates.  We leave it for future work.
\end{document}

>>>>

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $N^X$ and $N^Y$.  Here $N^X_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $N^Y_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.  For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses the data to produce two objects:
%
    \item $\hat q(y|x)$ -- a point estimate for the link parameter $q^*(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link parameter $q^*(y|x)$ with high probability
%
To produce these estimates for the link, we take a bayesian point of view.  It might seem sensible to start by specifying some prior beliefs about the link $q^*$.  However, the link is in many cases `nonidentifiable';  we describe this problem in detail in Appendix \ref{sec:casestudies}.  One consequence of nonidentifiability is that standard bayesian inferences would have no robustness to our precise choice of prior beliefs.  To avoid these problems, we focus on quantities which can be identified from the data.  In particular,
\begin{itemize}
    \item Let $p^*(x|\ell)$ denote the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$.  These probabilities can be estimated from the data, $N^X$.
    \item Let $h^*(y|\ell) \triangleq \sum_x p^*(x|\ell)q^*(y|x)$ denote the probability that a cell sampled with procedure $\ell$ will have morphological type $y$.  These probabilities can be estimated from the data, $N^Y$.
    \item Let $\Theta(p,h)=\{q:\ h(y|\ell) = \sum_x p(x|\ell)q(y|x)\}$ denote the set of links which are consistent with $p,h$.  Notice that $q^*\in\Theta(p^*,h^*)$, but there may be other links $q\in\Theta(p^*,h^*)$.  Any such link yields the same distribution on what we can observe, $N^X,N^Y$.  Therefore there is simply no way to use data to distinguish among them.  This is the essence of the nonidentifiability problem mentioned above.  So, instead of trying to estimate this inestimable quantity, we will focus on three other quantities which can be identified from data.  Let
    \begin{itemize}
        \item $q_{\mathrm{lo},x,y} \triangleq \min_{q\in \Theta(p^*,h^*)} q(y|x)$
        \item $q_{\mathrm{hi},x,y} \triangleq \max_{q\in \Theta(p^*,h^*)} q(y|x)$ 
        \item $q_{\mathrm{mid}} \triangleq \argmin_{q\in \Theta(p^*,h^*)} \sum_{x,y} q(y|x)^2$
    \end{itemize}
    Note that all three of these optimization problems have unique solutions and would be straightforward to solve if $p^*,h^*$ were known; optimization details can be found in Appendix \ref{sec:optproblems}.  Morever, even when $q^*$ is nonidentifiable, we can still be assured that $q_{\mathrm{lo}}, q_{\mathrm{hi}}$ are identifiable and $q_{\mathrm{lo},x,y} \leq q^*(y|x) \leq q_{\mathrm{hi},x,y}$.  We can also hope that $q_{\mathrm{mid}}$ will provide a single estimate which strikes a middle ground; in particular, $q_{\mathrm{mid}}$ is the link value which is closest to uniform among all links which are consistent with $p^*,h^*$.  
\end{itemize}  
To summarize: it may not be possible to estimate $q^*$ from data, but we can always estimate $p^*,h^*$.  We can also compute deterministic functions of $p^*,h^*$ which bound $q^*$, namely $q_{\mathrm{lo}},q_{\mathrm{hi}}$.  We take a Bayesian point of view in estimating $p^*,h^*$.  For prior beliefs, we assume a noninformative uniform prior $\mathbb{P}$, namely
\[
\mathbb{P}(p^*,h^*) \propto 1
\]
Following the bayesian philosophy, we then incorporate new knowledge by conditioning.  We have two important pieces of knowledge we would like to incorporate.  First, we have observed the data, $N^X,N^Y$.  Second, we know from the MLM assumption that there exists \emph{some} value $q^*$ which is consistent with $p^*,h^*$.   To make the second idea formal, let $D(h|h')$ denote the Kullback-Leibler divergence and $G(p,h) = \min_{h':\ \Theta(p,h')\neq \emptyset} D(h|h')$.  $G(p,h)=0$ is a numerical way to indicate that the MLM assumption can hold.  After incorporating this knowledge, we find ourselevs interested in the conditional distribution
\[
\mathbb{P}\left(p^*,h^* | N^X,N^Y,G(p^*,h^*)=0\right)
\]
We use this posterior to produce estimates for our quantities of interest:
    \item $\hat q$ is calculated using posterior expectation:
    \[
    \hat q \triangleq \mathbb{E}\left[q_{\mathrm{mid}} | N^X,N^Y,G(p^*,h^*)=0\right]
    \]
    %
    \item $C_{x,y}$ is calculated using credible intervals.  For each $x,y$, we define $C_{x,y}$ to be the largest interval such that 
    \begin{align*}
    \mathbb{P}\left(q_{\mathrm{lo},x,y} \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% \\
    \mathbb{P}\left(q_{\mathrm{hi},x,y} \in C_{x,y} | N^X,N^Y,G(p^*,h^*)=0\right)\geq 97.5\% 
    \end{align*}
Code to compute these objects is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

In practice, we were not able to find a way to compute the credible interval and posterior means exactly.  It seems difficult to even sample from the conditional distribution.  Common approaches to this type of problem involve Markov-Chain Monte Carlo and Variational Inference, but we were unable to make these approaches work in practice.  It seems nontrivial to work with the condition $G(p,h)=0$ that formalizes the MLM assumption.  We instead take a somewhat na\"ive approach.  We start by defining $\tilde P,\tilde H$ to carry the conditional distribution of $p^*,h^*$ given the observed data (but not conditioned on the fact that the MLM assumption holds).  We then simply project $\tilde H$ to the space of parameters which are consistent with the MLM assumption.  That is, we take $\hat H$ to be the solution of the minimization problem $\min_{h'}D(\tilde H|h')$, subject to the constraint that $E(\tilde P,h')=0$.  Optimization details can be found in Appendix \ref{sec:optproblems}.  We hope that the distribution of $\tilde P,\hat H$ will approximate the desired posterior.  We then use samples from this distribution to compute what we need.  We asymptotically expect that $\tilde P,\tilde H$ will nearly satisfy the MLM assumption in any case, so this approximation should not make a large difference.  However, we caution that the exact implications of the approximation are difficult to reason about.   In any case, it is a practical solution to a difficult problem.

@@

\section{Mathematical Results and Simulations}

@@

There are three desiderata we might hope the Markov Link Method estimators would achieve:
\begin{description}
    \item[Estimator convergence] As we obtain more samples, $\hat q(y|x)$ should converge to the true link $q^*(y|x)$ for each $x,y$.
    \item[Interval concentration] As we obtain more samples, the interval $C_{x,y}$ should get smaller. Asymptotically it should include nothing but the point $q^*(y|x)$.
    \item[Conservative coverage] $q^*(y|x) \in C_{x,y}$ with high probability.
\end{description}
However, due to certain identifiability issues, it is only actually possible to guarantee the last desideratum.  In many cases, the MLM achieves all three.  However, we can show that identifiability issues block any possibility of success in some cases; we detail this problem in Appendix \ref{sec:casestudies}.  The good news is that even though the Markov Link Method credible intervals may not concentrate around the truth, they do as well as possible.  In particular, they do appear to converge around the quantities we can estimate, namely the $q_\mathrm{lo},q_\mathrm{hi}$ bounds described in Section \ref{sec:mlm}.  A precise  statement of what we have been able to prove may be found in Theorem \ref{thm:mainthm} of Appendix \ref{sec:casestudies}.

The following two simulations give a good overview of the relevant issues:

@@

The MLM appears gives provable guarantees, provides reasonable values in simulation, and gives useful insight to real data.  Applied to neuronal cell-type data, the MLM clarified how two different cell-type classification systems are related.  We saw that some aspects of the two systems seem well-calibrated, but others we are less sure about.  The nature of the variability in the credible intervals suggested directions for experiments which would further refine our understanding.

@@

Another direction for future work would be to take a frequentist point of view.  An obvious direction would be to use profile likelihoods as statistics to define confidence intervals for the link parameters.  However, the distribution of these statistic is difficult to pin down.  Traditional techniques based on asymptotic normality may fail dramatically, because the true parameters may lie at the boundary of the parameter space.  In particular, there may be some $x,y$ for which $q^*(y|x)=0$.  Even approximate methods are difficult to apply because the parameter space can be quite high-dimensional.  Perhaps future work could uncover a way to overcome these challenges.

It is clear that good assumptions can help us get real insight for tough problems. Even if these assumptions are not sufficient to allow us to perfectly identify our object of interest, we can bound our uncertainty.  By probing this uncertainty carefully, we can learn what the data actually has to say and what experiments will help us learn more.

@@

Since it is easy to find consistent estimators for $\hat p,\hat h$, this theorem suggests we can use those estimators to get good estimates for our uncertainty about $q^*$.  In particular, subject to the conditions of the theorem, we have that the bounds of $C_{x,y}$ converge to the bounds of what is identifiable about $q^*$.  It is certainly a step in the right direction, but we emphasize that the theorem's conditions are nontrivial.  It is our opinion that the linear independence condition is fairly mild (if it is not met then subpopulations can simply be merged together).  However, the positivity condition is quite troubling.  In many cases of interest the true link $q^*$ has genuine zeros: pairs of measurements between the two tools which are fundamentally incompatible.  In this case such a theorem cannot be applied.

However, it may be that the above theorem's requirement ($q(x|y)>c>0$) is much stronger than is necessary.  A precise understanding of the discontinuity example above has remained elusive.  Better understanding could lead to more accurate estimates.  We leave it for future work.

\end{document}
